Stateful Delta Lenses

### 2018-02-14 papers/SDLenses/main.tex
Made some edits

Unfortunately, the lens laws only require that the \pput component of a lens
return the original source unmodified whenever the update to the view is a
no-op; they do not specify the behaviour of the \pput component when the update
to the view is non-trivial. Consequently, a lens may be valid even though it
does not propagate non-trivial updates made to the view back into the source in
the manner intended. For example, a small update to the view can result
in a large update to the source, such as when a lens updates records
positionally even though the source and view are order agnostic. This paper
introduces {\em stateful delta lenses} or SD-lenses as a solution to this
shortcoming of traditional lenses. SD-lenses are lenses augmented with extra
information that specifies what kind of updates are valid for a given lens, and
how these updates should be propagated by the lens, and hence enable the
programmer to specify more exactly the desired behaviour of a lens.
\end{abstract}

@@

Lenses are designed to perfom these kinds of computations. If $\ell$ is
a suitably defined lens for the bidirectional tranfsormation that we
have just described, then the \get component of the lens performs the
XML to ASCII transformation that produces an ASCII view from an XML source,
while the \pput component of the lens performs the ASCII to XML transformation
that folds back an updated ASCII view back into the original XML source.

Formally, Foster \cite{foster2007combinators,foster2009bidirectional} defined a lens $\ell :
S \Leftrightarrow V$ from a source $S$ to a view $V$ to be a triple of
functions $\ell.\get : S \longrightarrow V$, $\ell.\pput : S \longrightarrow S
\longrightarrow V$ and $\ell.\create : V \longrightarrow S$ satisfying the
following properties:
\begin{align*}
\ell.\get \; (\ell.\pput \; s \; v) &= v \tag{PUTGET}\\
\ell.\get \; (\ell.\create \; v) &= v \tag{CREATEGET}\\
\ell.\pput \; s \; (\ell.\get \; s) &= s \tag{GETPUT}
\end{align*}
The original definition of lenses in \cite{foster2007combinators} involved only the \get
and \pput functions, but a \create function was later added in
\cite{foster2009bidirectional} to handle the special case where we need to construct a
source from a view, but we have no source to use as the original. \sam{Perhaps
we should include this in the footnotes?}

@@

While these two laws give some guarantees on how the \get and \pput components
of a lens interact, they are not strong enough to guarantee that the
updates made to a view are indeed the updates reflected back into the source.
This is because the GETPUT law only says that the \pput component must return
the original source unmodified whenever the update to the view is a no-op; it
does not say what the \pput component should do with a modified view.
Consequently, a non-trivial update to a view could lead to an undesired update
to the source.

For example, suppose that a source $S$ consists of a pair of integers, and
that the view $V$ for $S$ consists of a single integer. Further, suppose that we
wish for the \get of any lens from $S$ to $V$ to simply project away
the first factor:
$\ell.\get \; (a, b) = b$. Suppose further that we wish to force the \pput
function of any lens $\ell$ defined on the data to restore the factor projected
away by the \get function: that is, we wish to enforce that for
all integers $a, b, c$, $\pi_1(\ell.\pput \; (a, b) \; c) = a$, where $\pi_1$ is
the projection on the first factor.

Now consider the lens $\ell_{bad}$ defined by $\ell_{bad}.\get \; (a, b) = b$
and
\begin{align*}
\ell_{bad}.\get \; (a, b) &= b\\
\ell_{bad}.\pput \; (a, b) \; c &= \begin{cases}
(a, c) & \text{if }b = c\\
(0, c) & \text{otherwise}\\
\end{cases}\\
\ell.\create \; b &= (0,b)
\end{align*}
Then $\ell_{bad}$ is a valid lens even though $\ell_{bad}$ does not
always restore the first factor. For example, observe that $\ell.\pput (1, 2) \;
3 = (0,3)$. The lens laws do not allow us to specify how non-trivial updates are
propagated, hence the \pput component can cause any update to the source, so
long as the \get component maps the updated source back to the updated view.

In this paper, we introduce {\em stateful delta lenses} or SD-lenses, which are
lenses augmented with extra information that specifies what kind of updates are
valid for a given lens, and how these updates should be propagated.

@@

Next we define a lens $\partial(S, V) : \partial S \Leftrightarrow \partial V$
which describes how updates should be propagated. For this example, we let
define $\partial(S,V)$ to simply be the \ccopy lens from $\partial S$ to
$\partial V$, keeping in mind that $\partial S = \partial V$:
\begin{align*}
(\partial(S, V)).\get \; (a, b) &= (a, b)\\
(\partial(S, V)).\pput \; (a, b) \; (c, d) &= (c, d)\\
(\partial(S, V)).\create \; (a, b) &= (a, b)
\end{align*}
To specify the intended behaviour of a lens $\ell$ from $S$ to $V$ we
require that
\begin{equation}\label{partialpreservesdelta}
(\partial (S, V)).\get \; (\delta_S(\ell.\pput \; s \; v, s)) = \delta_V(v,
\ell.\get \; s)
\end{equation}
This condition says that if we start with a source $s$, map $s$ to the view
to get $\ell.\get \; s$ and then update $\ell.\get \; s$ to get a new view $v$,
then the \get component of $\partial(S, V)$ applied to $\delta_S(\ell.\pput \; s
\; v, s)$ is equal to $\delta_V(v, \ell.\get \; s)$.

For example, observe that $\ell_{bad}$ does not satisfy this condition since
\begin{align}
(\partial (S, V)).\get \; (\delta_S(\ell_{bad}.\pput \; (1,2) \; 3, (1,2))) &=
(\partial (S, V)).\get \; (\delta_S((0,3), (1,2)))\\
&= (\partial (S, V)).\get \; (0,3)\\
&= (0,3)\\
&\neq (\epsilon, 3)\\
&= \delta_V(3, 2)\\
&= \delta_V(3, \ell_{bad}.\get \; (1, 2))
\end{align}
Consequently $\ell_{bad}$ is not a valid SD-lens. On the
other hand, the lens $\ell_{good} : S \Leftrightarrow V$ defined by
\begin{align*}
\ell_{good}.\get \; (a, b) &= b\\
\ell_{good}.\pput \; (a, b) \; c &= (a, c)\\
\ell_{good}.\create \; b &= (0, b)
\end{align*}
is indeed a SD-lens since the its \pput component always restores the first
factor.

Up to this point we have not yet explained the role of inclusion functions. Now
assume that we have valid lenses $\ell_1$ and $\ell_2$ from source $S_i$ to
view $V_i$ for $i \in  \{1, 2\}$. One of the standard lens combinators is the
conditional combinator $\ell_1 \sqcup \ell_2$ which uses $\ell_1$ for $s \in
S_1$ and $v \in V_1$ and $\ell_2$ for $s \in S_2$ and $v \in V_2$. Now assume
that $s_1 \in S_1$ and $s_2 \in S_2$. When computing the difference between
$s_1$ and $s_2$, we wish to say that $\delta_{S_1 \sqcup S_2}(s_1, s_2) = s_1$
since $s_1$ and $s_2$ are in disjoing sets and are hence incomparable, but $s_1$
is in $S_1$ rather than in $\partial S_1$! Hence for any source $S$ we need to
embed $s$ as an update inside $\partial S$ and similarly for views, hence the
need for inclusion functions $i_S : S \longrightarrow \partial S$ and $i_V : V
\longrightarrow \partial V$. For inclusion functions $i_S$ and $i_V$, we
require a law similar to \cref{partialpreservesdelta} to hold:
$$(\partial (S, V)).\get \; (i_{S}(s)) = i_{V}(\ell.\get \; s)$$
\section{Stateful Delta Lenses}
\begin{definition}
Let $(S, \partial S, \delta_S, i_S)$ and $(V, \partial V, \delta_V, i_V)$ be
difference structures. A stateful delta-lens (SD-Lens) $\ell : (S, \partial S,
\delta_S, i_S) \Leftrightarrow (V, \partial V, \delta_V, i_V)$ is a pair
$(\ell, \partial (S, V))$ such that for all $s \in S$ and $v \in V$,
\begin{enumerate}
  \item
  $\ell : S \Leftrightarrow V$ is a lens,
  \item
  $\partial(S, V) : \partial S \Leftrightarrow \partial V$ is a lens,
  \item
  $(\partial (S, V)).\get \; (\delta_S(\ell.\pput \; s \; v, s)) = \delta_V(v,
  \ell.\get \; s)$
  \item
  $(\partial (S, V)).\get \; (i_{S}(s)) = i_{V}(\ell.\get \; s)$
\end{enumerate}

>>>>

Unfortunately, the lens laws only require that the \pput component of a lens
return the original source unmodified whenever the update to the view is a
no-op; they do not specify the behaviour of the \pput component when the update
to the view is non-trivial. Consequently, a lens may be valid even though it
does not propagate non-trivial updates in the manner intended. For example, a
small update to the view can result in a large update to the source, such as
when a lens updates records positionally even though the source and view are
order agnostic. This paper introduces {\em stateful delta lenses} or SD-lenses
as a solution to this shortcoming of traditional lenses. SD-lenses are lenses
augmented with extra information that explicitly specifies the updates that can
be made for a given lens and how these updates should be propagated. SD-lenses
explicitly specify the intended behaviour of a lens hence enable programmers to
give much stronger guarantees even when the update to a view is non-trivial.
\end{abstract}

@@

Lenses are designed to perfom these kinds of bidirectinoal transformations. If
$\ell$ is a suitably defined lens for the tranfsormation that we
have just described, then the \get component of the lens performs the
XML to ASCII transformation that produces an ASCII view from an XML source,
while the \pput component of the lens performs the ASCII to XML transformation
that folds back an updated ASCII view back into the original XML source.

Formally, Foster \cite{foster2007combinators,foster2009bidirectional} defined a lens $\ell :
S \Leftrightarrow V$ from a source $S$ to a view $V$ to be a triple of
functions $\ell.\get : S \longrightarrow V$, $\ell.\pput : S \longrightarrow S
\longrightarrow V$ and $\ell.\create : V \longrightarrow S$ satisfying the
following properties:
\begin{align*}
\ell.\get \; (\ell.\pput \; s \; v) &= v \tag{PUTGET}\\
\ell.\get \; (\ell.\create \; v) &= v \tag{CREATEGET}\\
\ell.\pput \; s \; (\ell.\get \; s) &= s \tag{GETPUT}
\end{align*}
The original definition of lenses in \cite{foster2007combinators} involved only
the \get and \pput functions, but a \create function was later added in
\cite{foster2009bidirectional} to handle the special case where we need to
construct a source from a view, but we have no source to use as the original.

@@

While these two laws give some guarantees on how the \get, \pput and \create
components of a lens interact, they do not specify the behaviour of the \pput
component when the update to the view is non-trivial, because the GETPUT law
only says that the \pput component must return the original source unmodified
whenever the update to the view is a no-op Consequently, a non-trivial update
to a view could lead to an undesired update to the source.

For example, suppose that a source $S$ consists of a pair of integers, and
that the view $V$ for $S$ consists of a single integer. Further, suppose that we
wish for the \get component of any lens from $S$ to $V$ to simply project away
the first factor:that is for any $\ell : \mathbb{Z} \times \mathbb{Z}
\Leftrightarrow \mathbb{Z}$, $\ell.\get \; (a, b) = b$ for all integers $a, b$.
Suppose further that we wish to force the \pput function of any lens $\ell$
defined on the data to restore the factor projected away by the \get function:
that is, we wish to enforce that for all integers $a, b, c$, $\pi_1(\ell.\pput
\; (a, b) \; c) = a$, where $\pi_1$ is the projection on the first factor.

Now consider the lens $\ell_{bad}$ defined by
\begin{align*}
\ell_{bad}.\get \; (a, b) &= b\\
\ell_{bad}.\pput \; (a, b) \; c &= \begin{cases}
(a, c) & \text{if }b = c\\
(0, c) & \text{otherwise}\\
\end{cases}\\
\ell.\create \; b &= (0,b)
\end{align*}
Then $\ell_{bad}$ is a valid lens even though $\ell_{bad}$ does not
always restore the first factor. For example, observe that $\ell.\pput (1, 2) \;
3 = (0,3)$. The lens laws do not allow us to specify how non-trivial updates are
propagated, hence the \pput component can cause any update to the source, so
long as the \get component maps the updated source back to the updated view.

SD-lenses are lenses augmented with extra information that explicitly specifies
the updates that can be made for a given lens and how these updates should be
propagated.

@@

Next we define a lens $\partial(S, V) : \partial S \Leftrightarrow \partial V$
which describes how updates should be propagated. For this example, we
define $\partial(S,V)$ to simply be the \ccopy lens from $\partial S$ to
$\partial V$, keeping in mind that $\partial S = \partial V$:
\begin{align*}
(\partial(S, V)).\get \; (a, b) &= (a, b)\\
(\partial(S, V)).\pput \; (a, b) \; (c, d) &= (c, d)\\
(\partial(S, V)).\create \; (a, b) &= (a, b)
\end{align*}
To specify the intended behaviour of a lens $\ell$ from $S$ to $V$ we
require that
\begin{equation}\label{partialpreservesdelta}
(\partial (S, V)).\get \; (\delta_S(\ell.\pput \; s \; v, s)) = \delta_V(v,
\ell.\get \; s)
\end{equation}
This condition says that if we start with a source $s$, map $s$ to the view
to get $\ell.\get \; s$ and then update $\ell.\get \; s$ to get a new view $v$,
then the \get component of $\partial(S, V)$ applied to $\delta_S(\ell.\pput \; s
\; v, s)$ is equal to $\delta_V(v, \ell.\get \; s)$.

This condition enable us to rule out lenses which do not propagate updates
correctly. For example, observe that $\ell_{bad}$ does not satisfy
\cref{partialpreservesdelta} since
\begin{align*}
(\partial (S, V)).\get \; (\delta_S(\ell_{bad}.\pput \; (1,2) \; 3, (1,2))) &=
(\partial (S, V)).\get \; (\delta_S((0,3), (1,2)))\\
&= (\partial (S, V)).\get \; (0,3)\\
&= (0,3)\\
&\neq (\epsilon, 3)\\
&= \delta_V(3, 2)\\
&= \delta_V(3, \ell_{bad}.\get \; (1, 2))
\end{align*}
Consequently $\ell_{bad}$ is not a valid SD-lens. On the
other hand, the lens $\ell_{good} : S \Leftrightarrow V$ defined by
\begin{align*}
\ell_{good}.\get \; (a, b) &= b\\
\ell_{good}.\pput \; (a, b) \; c &= (a, c)\\
\ell_{good}.\create \; b &= (0, b)
\end{align*}
is indeed a valid SD-lens since the its \pput component always restores the
first factor.

Up to this point we have not yet explained the role of inclusion functions. Now
assume that we have valid lenses $\ell_1$ and $\ell_2$ from source $S_i$ to
view $V_i$ for $i \in  \{1, 2\}$. One of the standard lens combinators is the
conditional combinator $\ell_1 \sqcup \ell_2$ which uses $\ell_1$ for $s \in
S_1$ and $v \in V_1$ and $\ell_2$ for $s \in S_2$ and $v \in V_2$. Now assume
that $s_1 \in S_1$ and $s_2 \in S_2$. When computing the difference between
$s_1$ and $s_2$, we wish to say that $\delta_{S_1 \sqcup S_2}(s_1, s_2) = s_1$
since $s_1$ and $s_2$ are in disjoint sets and are hence incomparable, but $s_1$
is in $S_1$ rather than in $\partial S_1$ and hence is not an update! Hence for
any source $S$ we need to embed $s$ as an update inside $\partial S$ and
similarly for views, hence the need for inclusion functions $i_S : S
\longrightarrow \partial S$ and $i_V : V \longrightarrow \partial V$. For
inclusion functions $i_S$ and $i_V$, we require a law similar to
\cref{partialpreservesdelta} to hold: $$(\partial (S, V)).\get \; (i_{S}(s)) =
i_{V}(\ell.\get \; s)$$
\section{Stateful Delta Lenses}
\begin{definition}
Let $(S, \partial S, \delta_S, i_S)$ and $(V, \partial V, \delta_V, i_V)$ be
difference structures. A stateful delta-lens (SD-Lens) $\ell : (S, \partial S,
\delta_S, i_S) \Leftrightarrow (V, \partial V, \delta_V, i_V)$ is a pair
$(\ell, \partial (S, V))$ such that for all $s \in S$ and $v \in V$,
\begin{enumerate}
  \item
  $\ell : S \Leftrightarrow V$ is a lens,
  \item
  $\partial(S, V) : \partial S \Leftrightarrow \partial V$ is a lens,
  \item
  $(\partial (S, V)).\get \; (\delta_S(\ell.\pput \; s \; v, s)) = \delta_V(v,
  \ell.\get \; s)$
  \item
  $(\partial (S, V)).\get \; (i_{S}(s)) = i_{V}(\ell.\get \; s)$
\end{enumerate}
